Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-10-17 papers/termination/termination-esop/paper.tex
fixed soundness and complexity so that they don't refer to the synthesizer

@@

\section{Soundness, Completeness and Complexity}

Termination for bit-vector programs is known to be
PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
expected results for soundness and completeness of our algorithm for
deciding it.  Our strategy for proving soundness and completeness
is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
with \newC as a background logic) is a semi-decision
procedure.  Next, we will show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
We will then combine these results to show that the procedure as a whole
terminates with a valid proof for every input program.
%
%% \begin{theorem}\label{thm:synth-sound}
%% Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
%% $(P, x_0) \models \sigma$.
%% \end{theorem}

@@

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula {\bf [GT]} for any loop $L$ is valid
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{theorem}
 Algorithm~\ref{alg:termination} is sound and complete -- it terminates for all input loops $L$ with
 a correct termination verdict.
\end{theorem}

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is valid, and hence satisfiable.  Corollary~\ref{cor:cegis-newc}
 shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.

Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
$2^{bk}$ times where $k$ is the number of variables used by $L$ and
$b$ is the number of bits used to store each variable.
The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
computing $ L $.  So if
the CEGIS loop iterates $n$ times, the
runtime is bounded by
$$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.

>>>>

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable
 and the domain of inputs $X$ is finite,
 the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
 exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.
\end{theorem}

@@

\section{Soundness, Completeness and Complexity}
The second-order SAT solver we are using is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables~\cite{kalashnikov}. 
%% Termination for bit-vector programs is known to be
%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
%% expected results for soundness and completeness of our algorithm for
%% deciding it.  Our strategy for proving soundness and completeness
%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
%% with \newC as a background logic) is a semi-decision
%% procedure.
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our technique terminates with a valid proof for every input program.
%
%% \begin{theorem}\label{thm:synth-sound}
%% Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
%% $(P, x_0) \models \sigma$.
%% \end{theorem}

@@

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula {\bf [GT]} for any loop $L$ is a tautology
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{theorem}
Our technique %Algorithm~\ref{alg:termination} 
is sound and complete -- it terminates for all input loops $L$ with
 a correct termination verdict.
\end{theorem}

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.
 In~\cite{kalashnikov}, we show that the second-order SAT solver is semi-complete, and so is guaranteed to find a satisfying
 assignment for \lstinline!spec!. %, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}),
 otherwise $P_N$ is a non-termination proof (Theorem~\ref{thm:snt}).  Exactly one of these purported proofs
 will be valid, and since we can check each proof with a single call to a SAT solver we simply
 test both and discard the one that is invalid.
\end{proof}

%%---- Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.

%% Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
%% $2^{bk}$ times where $k$ is the number of variables used by $L$ and
%% $b$ is the number of bits used to store each variable.
%% The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
%% with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
%% computing $ L $.  So if
%% the CEGIS loop iterates $n$ times, the
%% runtime is bounded by
%% $$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
%% Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.
